---
title: "Linear regression: Abalone growth and pH example"
author: "Tadeh Peroomian"
date: "2026-05-15"
format: 
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

# Set up

```{r}
#| label: libraries-and-data

library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects) # generating model predictions
library(flextable) # creating tables

# abalone data from Hamilton et al. 2022
abalone <- read_xlsx(here("data", "Abalone IMTA_growth and pH.xlsx"))
```

# Abalone example

Data from Hamilton et al. 2022. "Integrated multi-trophic aquaculture mitigates the effects of ocean acidification: Seaweeds raise system pH and improve growth of juvenile abalone." https://doi.org/10.1016/j.aquaculture.2022.738571

## a. Questions and hypotheses

Question: How does pH predict abalone growth (measured in change in shell surface area per day, mm^2^ d^-1^)?  

H~0~: **pH has no effect on abalone shell growth**  

H~A~: **pH does have an effect on abalone shell growth**

## b. Cleaning

```{r}
#| label: abalone-data-cleaning

# creating new object from orgiinal data farm
abalone_clean <- abalone |> 
  #  cleaning names to remove spaces
  clean_names() |> 
  # selecting specific columns
  select(mean_p_h, change_in_area_mm2_d_1_25) |> 
  # renaming selected columns
  rename(mean_ph = mean_p_h, 
         change_in_area = change_in_area_mm2_d_1_25)
```

Don't forget to look at your data! Use `View(abalone_clean)` in the _Console_.

## c. Exploratory data visualization

```{r}
#| label: abalone-scatterplot
#| warning: false

# base layer: ggplot
ggplot(data = abalone_clean, # data comes from clean data set
       aes(x = mean_ph, # x-axis: pH, y-axis: change in shell surface area/day
           y = change_in_area)) +
  # second layer: points to represent observations
  geom_point(size = 4,
             stroke = 1,
             fill = "firebrick4",
             shape = 21)
```

Where does the missing value come from?  

**The missing value comes from one pair of tanks in which the pH sensor malfunctioned**  

Don't forget to turn your warnings off either in the individual code chunk or in the YAML at the top of the file.

## d. Abalone model

### Model fitting

```{r}
#| label: abalone-model

abalone_model <- lm(
  change_in_area ~ mean_ph, #  change in shell SA/day as a function of pH
  data = abalone_clean      # data frame: cleaned abalone data
)
```

### Diagnostics

```{r abalone-model-diagnostics}

par(mfrow = c(2, 2))   # displays plots in a 2x2 grid
plot(abalone_model)    #  displays diagnostic plots for model

```

### Model summary

```{r}
#| label: abalone-model-summary

# more information about the model
summary(abalone_model)

```

What is the slope estimate (with standard error?)  

$\beta_1$: 2.98 $\pm$ 0.76  

What is the intercept estimate (with standard error?)  

$\beta_0$: -18.50 $\pm$ 6.16

### Generating predictions

```{r}
#| label: abalone-model-predictions

# creating a new object called abalone_preds
abalone_preds <- ggpredict(
  abalone_model,      # model object
  terms = "mean_ph"   # predictor (in quotation marks)
)

# display the predictions
abalone_preds
```

Look at the column names using `colnames(abalone_preds)` in the Console.  

What are the column names?  

**insert your notes here**  

Look at the actual object (by clicking on it in Environment).  

Look at the class of the object using `class(abalone_preds)` in the Console.  

What is this type of object?  

**insert your notes here**  

View `abalone_preds` like a data frame. 

### Why generate model predictions?  

This dataset does not include any _actual_ observations of abalone growth at pH = 8.  

However, the model can tell us a _prediction_ for what abalone growth could be if pH = 8 based on the existing data.  

Figure out what abalone growth the model would _predict_ at pH = 8.

```{r}
#| label: model-predictions-8

# finding the model prediction at a specific value
ggpredict(
  abalone_model,      # model object
  terms = "mean_ph[8]"   # predictor (in quotation marks) and predictor value in brackets
)
```

How would you interpret this?  

**insert your response here**

### Visualizing model predictions and data

```{r}
#| label: abalone-model-final-figure

# base layer: ggplot
# using clean data frame
ggplot(data = abalone_clean,
       aes(x = mean_ph,
           y = change_in_area)) +
  # first layer: points representing abalones
  geom_point(size = 4,
             stroke = 1,
             fill = "firebrick4",
             shape = 21) +
  # second layer: ribbon representing confidence interval
  # using predictions data frame
  geom_ribbon(data = abalone_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.1) +
  # third layer: line representing model predictions
  # using predictions data frame
  geom_line(data = abalone_preds,
            aes(x = x,
                y = predicted)) +
  # axis labels
  labs(x = "Mean pH", 
       y = expression("Change in shell area ("*mm^{2}~d^-1*")")) +
  # theme
  theme_minimal()
```

### Creating a table with model coefficients, 95% confidence intervals, and more

**Note:** this function is from `flextable`.

```{r}
#| label: abalone-table

as_flextable(abalone_model)
```

**END OF ABALONE EXAMPLE**
